Field and experimental data together with computational

### 2022-03-08 manuscript_1.0/manuscript_word.Rmd
Updated manuscript tp the PLOS Comp. Biol. Updated figures with new figures.

---
title: Field and experimental data together with computational models
  reveal key cognitive mechanisms behind cleaner fish behaviour
author:
- Andrés E. Quiñones^[Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
  - andresqp@gmail.com]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
fig_caption: yes
fontsize: 12pt
header-includes:
- \usepackage{float}
- \usepackage{setspace}\doublespacing
- \linespread{2}
- \usepackage{lineno}
- \linenumbers
- \newcommand{\beginsupplement}{ \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
keep_tex: yes
linespread: 2
bibliography: Cleanerlearning.bib
style: M:\Zotero\styles\behavioral-ecology-and-sociobiology.csl
spacing: double
abstract: |
  While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings yield similar performances, and Lloyd Morgan’s cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish *Labroides dimidiatus*, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model.  We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance. 
keywords: learning, behaviour, cleaners, bayesian statisitics
---

@@

Often alternative cognitive mechanisms yield similar behaviour and/or cognitive
performances. This poses a problem for disentangling the mechanistic 
underpinnings of behaviour. This is particularly clear in research aimed at
discovering between species variation in *higher* cognitive abilities; or 
in other words, research on whether non-human animals show cognitive 
abilities believed to be uniquely human. For instance, when researchers 
try to find *mental time travel-like* behaviour, they usually come up with 
clever experiments to show the behaviour displayed requires inferences made 
through past events[@dally_Foodcaching_2006]. However, they often face the
challenge of alternative scenarios where simpler explanations, 
like classic associative learning, can bring about the 
observed behavioural outcome [@suddendorf_Evolution_2007]. Similarly, attempts 
to demonstrate the presence of *theory of mind* in non-human animals 
face objections justified by alternative mechanisms underpinning 
similar behavioural results [@heyes_Theory_1998]. 
Such controversies are usually settled by using the principle 
of parsimony and its cognitive version, 
Lloyd Morgan’s cannon, which states that the simpler explanation (mechanism)
should be accepted. Ideally, however, alternative hypotheses should be evaluated 
in light of their explanatory power. One potential dimension that could allow
the variation necessary among distinct cognitive mechanisms is environmental
variation. Controlled laboratory conditions often mask between and within 
species variation that arises among different environments. Thus, if 
predictions from the alternative mechanisms depend on environmental 
conditions, it is possible to assess the explanatory power 
of the different hypotheses by performing behavioural test in the context of
different environmental conditions.

From an evolutionary perspective, mechanisms are likely selected because 
of how they allow individuals to respond to environmental variation. 
For example, biological market theory predicts that the exchange rate of goods 
and/or services traded between cooperative partners adjusts to the 
law of supply and demand, when individuals have some degree of partner choice
[@noe_Biological_1995]. Supply and demand conditions, which typically 
depend on the abundance of the species involved, certainly vary in time 
and space. Therefore, natural selection should favour the ability 
to flexibly adjust decisions and behavioural output to 
current market conditions. Indeed, such adjustments have been repeatedly 
documented [@axen_Signalling_1996]. In animals, an obvious general candidate 
mechanism for the strategic adjustment is the cognitive machinery. However, 
it is not clear which cognitive mechanisms allow individuals to adjust their 
behaviour to the varying conditions. As adjustments to changes in 
market conditions have been documented in a great variety of species with 
highly variable brain anatomies, the question arises to what extent 
mechanisms beyond basic associative learning may be involved.

@@

Recent research on the cognitive tool kit needed to solve the ephemeral 
reward task has used a broad approach as proposed by Shettleworth 
[@shettleworth_Cognition_2009], who defined cognition as including 
all ways in which animals acquire information 
through the senses, process, retain and decide to act on it. 
Indeed, perception of relevant cues is of major importance 
for individual performance. For example, some animals find relevant 
information in the plates  [@wismer_Cuebased_2019] while others find 
it in the food [@pretot_Comparative_2021; @pretot_Comparing_2016]. 
However, identifying cues as salient is not the only challenge 
of the ephemeral reward task, as revealed by proximate learning models. 
Such models allow varying the cognitive tool kit and evaluating 
which minimal kit is necessary to solve 
the task at hand (e.g. [@dubois_Model_2021]). Applied to the ephemeral 
reward task, learning models showed that basic reinforcement learning does 
not suffice to solve the ephemeral reward task [@prat_Modelling_2022;
@quinones_Reinforcement_2019]. 
This is particularly so when models assume the more complex natural 
situation in which cleaner fish face not only resident-visitor pairs but also 
visitor-visitor and resident-resident pairs, as well a resident alone or 
a visitor alone. To be able to give visitors priority over residents, 
cleaners need to be able to assess a client’s value separately for the 
three possible scenarios (alone, paired with a fish with the 
same strategic option, paired with a fish with the alternative strategic option) 
[@quinones_Reinforcement_2019]. The ability to distinguish and value 
one stimulus differently alone from compound versions of it has been termed 
configurational learning, chunking, 
or segmentation (see references in [@prat_Modelling_2022]).

To solve the ephemeral reward task, cleaners need to account for 
the future consequences of current decisions. In the model by 
Quiñones *et al.* [@quinones_Reinforcement_2019], this could be achieved in 
two non-mutually exclusive ways: through low temporal discounting of 
future effects, also termed ‘chaining’ [@enquist_Power_2016]; and/or 
through perceiving a visitor client leaving as psychological punishment 
(i.e. as a negative reinforcer). Low temporal discounting is when 
individuals include in their valuation of an action the reward effects 
that this will have in the future. This is done by combining in a single 
valuation the reward obtained in the current time with all the reward that 
comes after, discounting for how far in the future reward 
is accrued. 'Chaining' the reward of these different time steps allows 
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations [@enquist_Power_2016]. Even though, 'chaining' 
can be readily implemented computationally in learning models 
[@enquist_Power_2016;@sutton_Reinforcement_2018], cognitively it seems to be 
a complex adaptation [@suddendorf_Evolution_2007]. On the other hand,
using client behaviour as a negative reinforcer is, in principle, 
easier to implement. Negative reinforcement is part of the ubiquitous 
learning mechanism termed operant conditioning [@thorndike_Animal_1898; 
@skinner_Behavior_1938]. Thus, the standard logic of 
Lloyd Morgan’s cannon demands that operant conditioning as the simpler 
explanation is to be accepted by default. Ideally, however, the two 
mechanisms should be evaluated in light of how well 
they explain the available data. Note that different fields interested in
cognition and decision making use different words to refer to negative 
reinforcers [@quinones_Reinforcement_2019;@sutton_Reinforcement_2018]. 
Here, for the sake of simplicity and clarity, we will use the word 
'penalty' to refer to this mechanism which includes a negative reinforcer.

Over the last decade, our team tested over a hundred wild-caught 
cleaner fish in the exact same paradigm of the ephemeral reward 
task [@salwiczek_Adult_2012;@wismer_Variation_2014;@triki_Decrease_2018;
@triki_Biological_2019;@triki_Brain_2020]. These fish often come from 
different reef locations around the study field site at Lizard Island, 
Great Barrier Reef in Australia. Further investigation of the local 
eco-sociological conditions revealed that cleaner and client 
fish population densities have a substantial impact on cleaner fish 
performance in the task. Cleaner fish from reef sites with relatively 
low densities were more likely to fail at solving the task [@triki_Biological_2019;
@triki_Decrease_2018; @wismer_Variation_2014]. The explanation for this pattern
is twofold. First, low cleaner fish population densities imply 
low supply and high demand for cleaning services. 
Under such conditions, there are fewer occasions under 
which failing to give priority to a visitor translates to an empty 
cleaning station. Second and also related to the state of the cleaning market,
visitor clients at low cleaner fish population densities 
are less likely to exert their partner choice options 
and are hence more likely to wait for inspection if not given cleaning 
priority [@triki_Brain_2020]. In the absence of partner choice 
cleaner fish should not give priority to visitors anymore.

@@

<!-- ### Overview -->

<!-- In this study, we combined empirical data from Triki *et. al*  -->
<!-- [@triki_Biological_2019;@triki_Brain_2020] and the computational  -->
<!-- model analysed in Quiñones *et al.*[@quinones_Reinforcement_2019]  -->
<!-- to test the relative validity of chaining versus operant conditioning as  -->
<!-- mechanisms to account for future consequences of cleaner fish desicion-making. -->
<!-- In order to compare the explanatory power of the different hypotheses,  -->
<!-- we estimated the parameter values of three versions of  -->
<!-- the model that best fit the empirical data.  -->
<!-- Particularly, we fitted the parameter that determines how important  -->
<!-- future reward is ($\gamma$), the magnitude of the negative reward  -->
<!-- ($\eta$) and a scaling constant that translates cleaner fish absolute  -->
<!-- to relative population abundances (see following sections for details).  -->
<!-- We obtained the model predictions using a set of stochastic individual-based -->
<!-- simulations. Thus, each parameter combination of the model yielded not one  -->
<!-- but a distribution of predictions. For the  -->
<!-- fitting procedure, we used the bayesian inference algorithm:  -->
<!-- Markov Chain Monte Carlo (MCMC). MCMC is a stochastic algorithm that allows -->
<!-- to sample from the posterior distribution of model parameter values. The posterior -->
<!-- distribution represents the most likely parameter values given the data.  -->
<!-- Due to the stochastic nature of MCMC, which repeatedly samples parameter values -->
<!-- and evaluates their likelihood, the distribution of likelihoods from the  -->
<!-- stochastic model is also explored. This makes MCMC and bayesian inference -->
<!-- particularly suitable for the computational models we evaluate. Bellow,  -->
<!-- we explain first the computational model and the data collection, and later  -->
<!-- how the MCMC was used to estimate parameter values. -->

@@

<!-- A further model analysis contrast the effect of these two  -->
<!-- parameters against variation in the relative abundance of cleaners  -->
<!-- and the two client types. Assuming certainty that visitors leave, the  -->
<!-- two parameters yield different  -->
<!-- predictions when relative client abundance is very high. Under such  -->
<!-- conditions, the analysis shows that under very high client (relative)  -->
<!-- abundance the presence of future rewards ($\gamma>0$) no longer favours  -->
<!-- the preference for the ephemeral clients, while the preference persists  -->
<!-- when the negative reward mediated by client behaviour ($\eta>0$) is  -->
<!-- included. Thus, the analysis suggests that future reward as an  -->
<!-- explanation for cleaner fish ability to prefer visitors is supported.  -->
<!-- At first sight, empirical data seem to support future rewards as  -->
<!-- the mechanism used by cleaners, as the statistical analysis  by  -->
<!-- Triki *et al.* [@triki_Biological_2019] data revealed that low  -->
<!-- cleaner-to-client ratios predict low preference for visitors and hence  -->
<!-- low performance in the ephemeral reward task. However, low cleaner-to-client  -->
<!-- ratios may also lead to low probabilities of visitors leaving, which would  -->
<!-- cause low preferences for visitors independently of the underlying  -->
<!-- mechanism [@quinones_Reinforcement_2019]. Furthermore, a combination  -->
<!-- of future and negative rewards, could still be yield a better explanation  -->
<!-- for the data than any single factor.  -->

@@

The empirical data were collected between 2010 and 2019 always during 
the austral winter months June to August from a total of five 
study reef sites (Corner Beach-CB, Horseshoe-HS, Mermaid Cove-MC, 
Northern Horseshoe-NHS, and The Crest-TC)  at Lizard Island 
($14.6682° S, 145.4604° E$), Great Barrier Reef, Australia. 
The data consist of three sets: fish censuses, field 
observations of cleaner-client interactions to quantify the probability 
of visitors leaving if made to wait, and the 
performance of wild-caught cleaner fish in the ephemeral reward test. 
In total, we have twelve site/year data sets for fish censuses and 
corresponding performance in the lab test. Thus, some sites were sampled more 
than once. To estimate the population density of cleaner fish 
and their clients at a given site in 
a given year, Triki *et al.* [@triki_Biological_2019] used a series of ten 
transects of $30m$ each. Observers swam along the transect lines placed on 
the reef and first counted the visible large-bodied adult fish 
(species with total length $TL \geq 10cm$) including cleaner fish 
on a width of $5m$, and then on the 
return individuals of small-bodied fish species ($TL < 10 cm$) \
on a width of $1 m$
(see Triki et al. 2018 for further details on fish censuses data collection). 
We then scaled the counts of cleaner fish, small-bodied, 
and  large-bodied  clients fish densities per $100 m^2$. 
From the study by Triki *et al* [@triki_Biological_2019] 
cleaner fish and large-bodied client populations densities were 
highly correlated, and only the former 
was hence used in the analyses as representative of both measures.

@@

(ref:post) Posterior distributions for parameter values $\gamma$ and $\eta$ for 
the three models ("Full model", "chaining only" and "penalty only"). We show 
the kernel density estimates, below the mode (black dot) and the 65% 
(light blue shade) and 95% (grey shade)  highest posterior density interval 
for the two parameters. On the top, panels a and b show posterior 
distributions for a full model, including chaining ($\gamma$) and penalty 
($\eta$). Panel c, shows the $\gamma$ estimate from model with only chaining. 
Panel d shows the $\\eta$ estimate from a model with only penalty. Panel e, 
shows a measure of fit for all models, namely the distribution of pseudo-$R^2$ 
obtained from sampling the posterior distribution of parameter values.

```{r post, fig.cap='(ref:post)',out.width='100%',fig.height=6}
library(data.table)
library(coda)
library(ggplot2)
library(cowplot)
library(ggdist)
library(ggpubr)
source(here("../R_files/posPlots.R"))
source(here("loadData.R"))
library("jsonlite")

# Name of the folders containing the MCMC runs
# scen1<-"MCMCclean_gam_nRew_sca_"
# scen2<-"MCMCclean_gam_sca_"
# scen3<-"MCMCclean_Nrew_sca_"

scen1<-"MCMCclean_gam_Nrew_sca_old_"
scen2<-"MCMCclean_gam_sca_old_"
scen3<-"MCMCclean_Nrew_sca_old_"

# NLables names for the three scenarios
labels.Scen<-c("both","gam","Nrew")
labelsPlot.Scen<-c("Full Model","Chaining","Penalty")

## Load files --------------------------------------------------------------

# Defaults for MCMC analysis
burn.in<-1000
thinning<-100

MCMCdata1<-loadMCMCrun(scen1,thinning = thinning,burn.in = burn.in)
MCMCdata2<-loadMCMCrun(scen2,thinning = thinning,burn.in = burn.in)
MCMCdata3<-loadMCMCrun(scen3,thinning = thinning,burn.in = burn.in)

# Create list with the data, convenient to use CODA for diagnostics
mcmcList.1<-mcmc.list(do.call(list,lapply(unique(MCMCdata1$seed), function(repli){
  rundata<-MCMCdata1[seed==repli]
  mcmcRun<-mcmc(rundata[,.(gamma,negReward,scaleConst)],thin = thinning)#
  return(mcmcRun)
})))

mcmcList.2<-mcmc.list(do.call(list,lapply(unique(MCMCdata2$seed), function(repli){
  rundata<-MCMCdata2[seed==repli]
  mcmcRun<-mcmc(rundata[,.(gamma,scaleConst)],thin = thinning)
  return(mcmcRun)
})))

mcmcList.3<-mcmc.list(do.call(list,lapply(unique(MCMCdata3$seed), function(repli){
  rundata<-MCMCdata3[seed==repli]
  mcmcRun<-mcmc(rundata[,.(negReward,scaleConst)],thin = thinning)#
  return(mcmcRun)
})))

## Density plots with gg -------------------------------------------------------

## Define the boundaries for credible intervals using High Density Interval

intervals<-c(0.66,0.95)

cuts.1<-lapply(MCMCdata1[,.(gamma,negReward,scaleConst)],
             function(x){
               c(mode_hdci(x,.width = intervals[c(2,1)])$ymin,
                 mode_hdci(x,.width = intervals)$ymax)
})

cuts.2<-lapply(MCMCdata2[,.(gamma,scaleConst)],
               function(x){
                 c(mode_hdci(x,.width = intervals[c(2,1)])$ymin,
                   mode_hdci(x,.width = intervals)$ymax)
               })

cuts.3<-lapply(MCMCdata3[,.(negReward,scaleConst)],
               function(x){
                 c(mode_hdci(x,.width = intervals[c(2,1)])$ymin,
                   mode_hdci(x,.width = intervals)$ymax)
               })

# get all the likelihoods from the posterior for all models
loglikehoods.all<-data.table(lglikelihood=c(MCMCdata1$fit,MCMCdata2$fit,
                                             MCMCdata3$fit),
                              model=c(rep(labels.Scen[1],length(MCMCdata1$fit)),
                                      rep(labels.Scen[2],length(MCMCdata2$fit)),
                                      rep(labels.Scen[3],length(MCMCdata3$fit))
                                      ))

# Read field data to calculate null log-likelihood
fieldData<-fread(here("Data","data_cleaner_abs_threa1.5.txt"))

nullLikehood<-sum(dbinom(x=fieldData[,score_visitor],
                         size = 20,prob = 0.5,log = TRUE))

# calculate pseudo R^2
loglikehoods.all[,Rsqrd:=1-lglikelihood/nullLikehood]

panel.rsqr.all<-
  ggplot(data=loglikehoods.all[is.finite(lglikelihood)],
                  aes(y=model,x=Rsqrd,fill=model))+
  stat_halfeye(alpha=0.5,point_size=3)+
  theme_classic()+xlim(-0.5,0.25)+
  scale_fill_manual(values=c("#1b9e77","#d95f02","#7570b3"),
  name = "", labels = labelsPlot.Scen)+
  ylab("")+
  scale_y_discrete(labels=labelsPlot.Scen)+
  geom_vline(xintercept = 0,color="black",size=0.5)+
  theme(legend.position = 'none')
          # c(.32, .65)
        # ,legend.key.size = unit(0.22,'cm'),
        # axis.text.y = element_blank(),
        # legend.text = element_text(size=7))

gam.both.post<-ggplot(data=MCMCdata1,aes(x=gamma)) +
  stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.1$gamma))),
               point_interval = mode_hdi, .width = c(.66, .95),
               point_size=3,show.legend=FALSE) +
  labs(title=labelsPlot.Scen[1],
       subtitle = expression(gamma),x="",y="")+
  theme_classic()+
  scale_fill_manual(values=c("gray85","skyblue","gray85"))

nrew.both.post<-ggplot(data=MCMCdata1,aes(x=negReward)) +
  stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.1$negReward))),
               point_interval = mode_hdi,
               .width = c(.66, .95),point_size=3,
               show.legend=FALSE)+theme_classic()+
  scale_fill_manual(values=c("gray85","skyblue","gray85"))+
  labs(title="",subtitle = expression(eta),x="",y="")+
  xlim(-2,2)
gam.gam.post<-ggplot(data=MCMCdata2,aes(x=gamma)) +
  stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.2$gamma))),
               point_interval = mode_hdi, .width = c(.66, .95),
               point_size=3,
               show.legend=FALSE) +
  scale_fill_manual(values=c("gray85","skyblue","gray85"))+
  labs(title=labelsPlot.Scen[2],
       subtitle = expression(gamma),x="",y="")+
  theme_classic()
nrew.nrew.post<-ggplot(data=MCMCdata3,aes(x=negReward)) +
  stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.3$negReward))),
               point_interval = mode_hdi, .width = c(.66, .95),
               point_size=3,
               show.legend=FALSE) +
  scale_fill_manual(values=c("gray85","skyblue","gray85"))+
  labs(title=labelsPlot.Scen[3],
       subtitle = expression(gamma),x="",y="")+
  theme_classic()+xlim(-2,2)

first.col<-plot_grid(nrow=3,align = "v",byrow = TRUE,
          gam.both.post,gam.gam.post,nrew.nrew.post,
          labels=c('a','c','d'))
second.col<-plot_grid(nrow = 2,align = "v",byrow = TRUE,
          nrew.both.post,panel.rsqr.all,rel_heights = c(1,2),
          labels = c('b','e'))
plot_grid(ncol = 2,first.col,second.col)

@@

(ref:rawdata) Relation between the response variable used in this study and the 
criteria used in previous studies to assess performance in the ephemeral reward 
task. In the x axis, we classified the performance of cleaner fish according to 
whether they developed a preference for the visitor in the initial round, 
in the initial and reversal, or none of them. In the y axis, we add the choices
of two experimental sessions: panel on the left uses one session from the initial 
round and one from the reversal round when possible 
(as described in the main text); panel on the right uses two sessions from the 
initial round for all fish.

```{r rawdata,fig.cap='(ref:rawdata)'}
plot_grid(initANDRev,init,ncol = 2)

>>>>

---
title: Field and experimental data together with computational 
  models reveal how cleaner fish adjust decisions  in a biological market
author:
- Andrés E. Quiñones^[Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
  - andresqp@gmail.com]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
fig_caption: yes
fontsize: 12pt
header-includes:
- \usepackage{float}
- \usepackage{setspace}\doublespacing
- \linespread{2}
- \usepackage{lineno}
- \linenumbers
- \newcommand{\beginsupplement}{ \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
keep_tex: yes
linespread: 2
bibliography: Cleanerlearning.bib
style: M:\Zotero\styles\behavioral-ecology-and-sociobiology.csl
spacing: double
abstract: |
  While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings yield similar performances, and Lloyd Morgan’s cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. We illustrate this point by fitting computational models to experimental data on performance by wild-caught cleaner fish *Labroides dimidiatus* in an ephemeral reward task, as well as cleaner and client fish densities from the locations of capture. Using Bayesian statistics to fit the model parameters to performance data revealed that cleaner fish most likely estimate future consequences of an action, while it appears unlikely that the removal of the ephemeral reward acts as psychological punishment (negative reinforcement). Incorporating future consequences also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement mechanism. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinnings of cognitive performance. 
author_summary: |
  Performance in behavioural experiments is often used to assess the cognitive abilities of animals. However, animals can get to the same outcome in alternative ways. Thus, the outcome of the experiments does not tell us how animals achieve their performance.  In order to overcome this limitation, we used a set of computational models, which provide predictions on how alternative cognitive mechanisms perform in the face of varying environmental conditions, and compare their predictions to performance data from individuals that experienced different conditions throughout their life. Our study system is the cleaner fish *Labroides dimidiatus*, a coral reef fish that feeds off ectoparasites and dead tissue from the skin of other reef fishes. Cleaner fish often have to choose among alternative clients seeking their cleaning service, and the optimal choice is often hard to achieve. Our performance experiments mimic the natural choices the cleaner fish make. The combination of experiments and computational models points to cleaner fish abilities to account for the future effect of their choices by estimating the long term expected value of those choices. Estimating long term value is a mechanism involved in human foresight.
keywords: learning, behaviour, cleaners, bayesian statisitics
---

@@

Often alternative cognitive mechanisms yield similar behaviour and/or cognitive
performances. This poses a problem for disentangling the mechanistic 
underpinnings of behaviour. This is particularly clear in research aimed at
discovering between species variation in *higher* cognitive abilities; or 
in other words, research on whether non-human animals show cognitive 
abilities believed to be uniquely human. For instance, when researchers 
try to find *mental time travel-like* behaviour, they usually come up with 
clever experiments to show the behaviour displayed requires inferences made 
through past events[@dally_Foodcaching_2006]. However, they often face the
challenge of alternative scenarios where simpler explanations, 
like classic associative learning, can bring about the 
observed behavioural outcome [@suddendorf_Evolution_2007]. Similarly, attempts 
to demonstrate the presence of *theory of mind* in non-human animals 
face objections justified by alternative mechanisms underpinning 
similar behavioural results [@heyes_Theory_1998]. 
Such controversies are usually settled by using the principle 
of parsimony and its cognitive version, 
Lloyd Morgan’s cannon, which states that the simpler explanation (mechanism)
should be accepted. Ideally, however, alternative hypotheses should be evaluated 
in light of their explanatory power. One potential dimension that could allow
the variation necessary among distinct cognitive mechanisms is environmental
variation. Controlled laboratory conditions often mask between and within 
species variation that arises among different environments. Thus, if 
predictions from the alternative mechanisms depend on environmental 
conditions, it is possible to assess their relative explanatory power 
by performing cognitive test in the context of
different environmental conditions.

From an evolutionary perspective, mechanisms are likely selected because 
of how they allow individuals to respond to environmental variation. 
For example, biological market theory predicts that the exchange rate of goods 
and/or services traded between cooperative partners adjusts to the 
law of supply and demand, when individuals have some degree of partner choice
[@noe_Biological_1995]. Supply and demand conditions, which typically 
depend on the abundance of the species involved, certainly vary in time 
and space. Therefore, natural selection should favour the ability 
to flexibly adjust decisions and behavioural output to 
current market conditions. Indeed, such adjustments have been repeatedly 
documented [@axen_Signalling_1996]. In animals, an obvious general candidate 
mechanism for the strategic adjustment is the cognitive machinery. However, 
it is not clear which cognitive mechanisms allow individuals to adjust their 
behaviour to the varying conditions. More specifically the question arises 
to what extent mechanisms beyond basic associative learning may be involved.

@@

Recent research on the cognitive abilities needed to solve the ephemeral 
reward task has used a broad approach as proposed by Shettleworth 
[@shettleworth_Cognition_2009], who defined cognition as including 
all ways in which animals acquire information 
through the senses, process, retain and decide to act on it. 
Indeed, perception of relevant cues is of major importance 
for individual performance. For example, some animals find relevant 
information in the plates  [@wismer_Cuebased_2019] while others find 
it in the food [@pretot_Comparative_2021; @pretot_Comparing_2016]. 
However, identifying cues as salient is not the only challenge 
of the ephemeral reward task, as revealed by proximate learning models. 
Such models allow varying the cognitive tool kit and evaluating 
which minimal kit is necessary to solve 
the task at hand (e.g. [@dubois_Model_2021]). Applied to the ephemeral 
reward task, learning models showed that basic reinforcement learning does 
not suffice to solve the ephemeral reward task [@prat_Modelling_2022;
@quinones_Reinforcement_2019]. 
This is particularly so when models assume the more complex natural 
situation in which cleaner fish face not only resident-visitor pairs but also 
visitor-visitor and resident-resident pairs, as well a resident alone or 
a visitor alone. To be able to give visitors priority over residents, 
cleaners need to be able to assess a client’s value separately for the 
three possible scenarios (alone, paired with a fish with the 
same strategic option, paired with a fish with the alternative strategic option) 
[@quinones_Reinforcement_2019]. The ability to distinguish and value 
one stimulus differently alone from compound versions of it has been termed 
configurational learning, chunking, 
or segmentation (see references in [@prat_Modelling_2022]).

In addition to configurational learning, cleaners also need to account for 
the future consequences of current decisions. In the model by 
Quiñones *et al.* [@quinones_Reinforcement_2019], this could be achieved in 
two non-mutually exclusive ways: through low temporal discounting of 
future effects, also termed ‘chaining’ [@enquist_Power_2016]; and/or 
through perceiving a visitor client leaving as psychological punishment 
(i.e. as a negative reinforcer). Low temporal discounting is when 
individuals include in their valuation of an action the reward effects 
that this will have in the future. This is done by combining in a single 
valuation the reward obtained in the current time with all the reward that 
comes after, discounting for how far in the future reward 
is accrued. 'Chaining' the reward of these different time steps allows 
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations [@enquist_Power_2016]. Even though, 'chaining' 
can be readily implemented computationally in learning models 
[@enquist_Power_2016;@sutton_Reinforcement_2018], cognitively it seems to be 
a complex adaptation [@suddendorf_Evolution_2007]. On the other hand,
using client behaviour as a negative reinforcer is, in principle, 
easier to implement. Negative reinforcement is part of the ubiquitous 
learning mechanism termed operant conditioning [@thorndike_Animal_1898; 
@skinner_Behavior_1938]. Thus, the standard logic of 
Lloyd Morgan’s cannon demands that operant conditioning as the simpler 
explanation is to be accepted by default. Ideally, however, the two 
mechanisms should be evaluated in light of how well 
they explain the available data. Note that different fields interested in
cognition and decision making use different words to refer to negative 
reinforcers [@quinones_Reinforcement_2019;@sutton_Reinforcement_2018]. 
Here, for the sake of simplicity and clarity, we will use the word 
'penalty' to refer to this mechanism which includes a negative reinforcer.

Over the last decade, our team tested over a hundred wild-caught 
cleaner fish in the exact same paradigm of the ephemeral reward 
task [@salwiczek_Adult_2012;@wismer_Variation_2014;@triki_Decrease_2018;
@triki_Biological_2019;@triki_Brain_2020]. These fish often come from 
different reef locations around the study field site at Lizard Island, 
Great Barrier Reef in Australia. Further investigation of the local 
eco-sociological conditions revealed that cleaner and client 
fish population densities have a substantial impact on cleaner fish 
performance in the task. Cleaner fish from reef sites with relatively 
low densities were more likely to fail at solving the task [@triki_Biological_2019;
@triki_Decrease_2018; @wismer_Variation_2014]. The explanation for this pattern
is twofold. First, low cleaner fish population densities imply 
low supply and high demand for cleaning services. 
Under such conditions, there are fewer occasions under 
which failing to give priority to a visitor translates to an empty 
cleaning station. Second and also related to the state of the cleaning market,
visitor clients at low cleaner fish population densities 
are less likely to exert their partner choice options 
and are hence more likely to wait for inspection if not given cleaning 
priority [@triki_Brain_2020]. In the absence of partner choice 
cleaner fish should not give priority to visitors any more.

@@

@@

@@

The empirical data were collected between 2010 and 2019 always during 
the austral winter months June to August from a total of five 
study reef sites (Corner Beach-CB, Horseshoe-HS, Mermaid Cove-MC, 
Northern Horseshoe-NHS, and The Crest-TC)  at Lizard Island 
($14.6682^\circ S, 145.4604^\circ E$), Great Barrier Reef, Australia. 
The data consist of three sets: fish censuses, field 
observations of cleaner-client interactions to quantify the probability 
of visitors leaving if made to wait, and the 
performance of wild-caught cleaner fish in the ephemeral reward test. 
In total, we have twelve site/year data sets for fish censuses and 
corresponding performance in the lab test. Thus, some sites were sampled more 
than once. To estimate the population density of cleaner fish 
and their clients at a given site in 
a given year, Triki *et al.* [@triki_Biological_2019] used a series of ten 
transects of $30m$ each. Observers swam along the transect lines placed on 
the reef and first counted the visible large-bodied adult fish 
(species with total length $TL \geq 10cm$) including cleaner fish 
on a width of $5m$, and then on the 
return individuals of small-bodied fish species ($TL < 10 cm$) \
on a width of $1 m$
(see Triki et al. 2018 for further details on fish censuses data collection). 
We then scaled the counts of cleaner fish, small-bodied, 
and  large-bodied  clients fish densities per $100 m^2$. 
From the study by Triki *et al* [@triki_Biological_2019] 
cleaner fish and large-bodied client populations densities were 
highly correlated, and only the former 
was hence used in the analyses as representative of both measures.

@@

<!-- (ref:post) Posterior distributions for parameter values $\gamma$ and $\eta$ for  -->
<!-- the three models ("Full model", "chaining only" and "penalty only"). We show  -->
<!-- the kernel density estimates, below the mode (black dot) and the 65%  -->
<!-- (light blue shade) and 95% (grey shade)  highest posterior density interval  -->
<!-- for the two parameters. On the top, panels a and b show posterior  -->
<!-- distributions for a full model, including chaining ($\gamma$) and penalty  -->
<!-- ($\eta$). Panel c, shows the $\gamma$ estimate from model with only chaining.  -->
<!-- Panel d shows the $\eta$ estimate from a model with only penalty. Panel e,  -->
<!-- shows a measure of fit for all models, namely the distribution of pseudo-$R^2$  -->
<!-- obtained from sampling the posterior distribution of parameter values. -->

```{r post, fig.cap='Posterior distributions for parameter values $\\gamma$ and $\\eta$ for the three models (Full model, chaining only and penalty only). We show the kernel density estimates, below the mode (black dot) and the 65 (light blue shade) and 95\\% (grey shade)  highest posterior density interval for the two parameters. On the top, panels a and b show posterior distributions for a full model, including chaining ($\\gamma$) and penalty ($\\eta$). Panel c, shows the $\\gamma$ estimate from model with only chaining. Panel d shows the $\\eta$ estimate from a model with only penalty. Panel e, shows a measure of fit for all models, namely the distribution of pseudo-$R^2$ obtained from sampling the posterior distribution of parameter values.',out.width='100%',fig.height=6,cache=FALSE}

source(here("analysisMCMC.R"))
first.col<-plot_grid(nrow=3,align = "v",byrow = TRUE,
          gam.both.post,gam.gam.post,nrew.nrew.post,
          labels=c('a','c','d'))
second.col<-plot_grid(nrow = 2,align = "v",byrow = TRUE,
          nrew.both.post,panel.rsqr.all,rel_heights = c(1,2),
          labels = c('b','e'))
plot_grid(ncol = 2,first.col,second.col)

@@

<!-- (ref:rawdata)  -->

```{r rawdata,fig.cap='Relation between the response variable used in this study and the criteria used in previous studies to assess performance in the ephemeral reward task. In the x axis, we classified the performance of cleaner fish according to whether they developed a preference for the visitor in the initial round, in the initial and reversal, or none of them. In the y axis, we add the choices of two experimental sessions: panel on the left uses one session from the initial round and one from the reversal round when possible (as described in the main text); panel on the right uses two sessions from the initial round for all fish.',out.width='100%',fig.height=5,cache=FALSE}
plot_grid(initANDRev,init,ncol = 2)
